---
title: "Mammal Temperature Data Wrangling"
output: html_document
---

# Install & load packages  
As always, we begin with the `install_packages()` (if needed) and `library()` commands to install and load individual packages, respectively.
```{r, message = FALSE}
library(tidyverse)
library(lubridate)
library(hms)
```

# Read in raw **mammal** data
```{r}
path <- "./raw_mammals/"

files <- list.files(path=path, pattern="*.csv")

for(file in files){
  perpos <- which(strsplit(file, "")[[1]]==".")
  assign(
    gsub(" ","",substr(file, 1, perpos-1)),
    read_csv(paste(path,file,sep=""), guess_max = 10000))
}
```

# Read in raw **temperature** data
```{r}
path <- "./raw_temp/"

files <- list.files(path=path, pattern="*.csv")

for(file in files){
  perpos <- which(strsplit(file, "")[[1]]==".")
  assign(
    gsub(" ","",substr(file, 1, perpos-1)),
    read_csv(paste(path,file,sep=""), guess_max = 500000))
}
```

# Wrangle temperature data
We always want to keep our **raw** data **raw**, so our next step will be to save a copy of our temperature data data as a different data object. Also for simplicity, we'll call it "temp" (shorter to type!)

First, we'll remove 'NA' values in the tempSingleMean column
```{r}
temp <- SAAT_30min %>% 
  filter(!is.na(tempSingleMean))
```

Next, we want to simplify our temperature data to at least a daily value. To do this, we will first use a `mutate()` function to create separate columns for date and time (rather than combined datetime). Next, we `select()` the columns we want to keep (including our new date and time columns)

To know which columns we might want to `select()`, it's helpful to have R re-print the column names for us.
```{r}
names(temp)
```

**Super-important note!** Make sure you include any data quality columns in your initial `select()` function! (in this file, the column is called **finalQF**)
```{r}
temp_sub <- temp %>% 
  mutate(date = as.Date(startDateTime),
         month = month(date),
         year = year(date),
         time = as_hms(startDateTime)) %>% 
  select(siteID, year, month, date, time, tempSingleMean:tempSingleMaximum, finalQF)

View(temp_sub)
```

We should only retain temperature values that do **not** have a failing quality flag (so our filter criterion would be: finalQF != 1 *or* finalQF == 0)
```{r}
temp_sub <- temp_sub %>% 
  filter(finalQF != 1)
```

Now, we can group and summarize our data to get (in this example) monthly temperature values
```{r}
monthly_temp <- temp_sub  %>% 
  group_by(siteID, year, month) %>% 
  summarize(monthlyMean = mean(tempSingleMean)) %>% 
  arrange(siteID, year, month) %>% 
  mutate(date = make_date(year, month))

View(monthly_temp)
```

If you want to save a copy of your *subset* data file, you can run the following command:
```{r}
monthly_temp %>% 
  write_csv('./cleaned_data/monthly_temp.csv')
```

# Quickly visualize cleaned up temperature data
It's always a good idea to take a quick look at your data to make sure nothing looks *off* (e.g., major outliers, etc)
```{r}
ggplot(monthly_temp, aes(x = date,
                     y = monthlyMean, 
                     col = siteID)) +
  geom_point() +
  theme_minimal()
```

# Clean mammal data
Our next step will be to save a copy of our raw mammal trapping data as a different data object. We'll call it "mam_raw"
```{r}
mam_raw <- mam_pertrapnight
```

Now we can use some of the same `dplyr` functions to begin filtering, organizing, and exploring the data
```{r}
mam <- mam_raw %>% 
  filter(!is.na(tagID)) %>% 
  select(siteID, plotID, collectDate, tagID:weight, dataQF) %>% 
  mutate(date = as.Date(collectDate),
         month = month(date),
         year = year(date))
```

If you want to save a copy of your *subset* data file, you can run the following command:
```{r}
mam %>% 
  write_csv('./cleaned_data/mammals_cleaned.csv')
```

Tally how many individuals per species were seen each month of sampling
```{r}
monthly_mam <- mam %>% 
  group_by(siteID, year, month, taxonID) %>% 
  tally()

print(monthly_mam)
```

Calculate monthly alpha diversity for each site as the number of distinct taxa seen per sampling month
```{r}
alpha <- monthly_mam %>% 
  group_by(siteID, year, month) %>% 
  summarize(alpha = n_distinct(taxonID))

print(alpha)
```
Quick plot of our alpha diversity data across sites:
```{r}
ggplot(alpha, aes(x = siteID,
                  y = alpha)) +
  geom_boxplot() +
  theme_minimal()
```

If you want to save a copy of any of your calculated values, you can do that too! (Here, we'll save a copy of the calculated alpha diversity)
```{r}
alpha %>% 
  write_csv('./cleaned_data/alpha_diversity.csv')
```

# Join monthly temperature and mammal data
We can also have R combine different data objects based on columns that they have in common. For our temperature and mammal data, they both share columns of siteID, year, and month 
```{r}
combined <- full_join(monthly_temp, alpha)
```

To save a copy of *this* output, you can again run a `write_csv()` command:
```{r}
combined %>% 
  write_csv('./cleaned_data/mammal_temp_combined.csv')
```

Finally, as a "teaser" for Thursday/next week, we can take a quick look at what the graph of monthly temperature vs. alpha diversity looks like:
```{r}
ggplot(combined, aes(x = monthlyMean,
                     y = alpha, 
                     col= siteID)) +
  geom_point() + 
  facet_grid(siteID~.) +
  theme_minimal()
```

